### 2025-03-12 paper/quarto_ecml/paper.tex
Review counterfactual training up to 3.4

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CEs,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this would comply with the definition of
plausibility if 40-year-old individuals were in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
(directly) mutable feature. For our intents and purposes, counterfactual
training aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.~consistent
with the data) and actionable. Formally, we define these properties as
follows:

@@

The definitions of faithfulness and plausibility in
Definition~\ref{def-explainability} are the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularises the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
different components can be governed by adjusting the strengths of the
penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

\subsubsection{Directly Inducing Explainability through Contrastive
Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
well at both tasks by directly maximizing the joint log-likelihood
factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of SGLD, we propose to use counterfactual explainers to generate
counterfactuals of observed training samples. Specifically, we have:

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
a mature counterfactual for a randomly sampled factual from a non-target
class generated through a gradient-based counterfactual generator for at
most \(T\) iterations. We define mature counterfactuals as those that
have either exhausted \(T\) or reached convergence in terms of the
pre-determined decision threshold earlier.

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
\cite{du2019implicit}. As the generated counterfactuals get more
plausible (Definition~\ref{def-explainability}) over the course of
training, these two opposing effects gradually balance each out
\cite{lippe2024uvadlc}.

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. Typically, these methods facilitate the imposition of domain
and mutability constraints, for example. In principle, any existing
approach for generating counterfactual explanations is viable, so long
as it does not violate the faithfulness condition. Like JEMs
\cite{murphy2022probabilistic}, counterfactual training can be
considered as a form of contrastive representation learning.

\subsubsection{Indirectly Inducing Explainability through Adversarial
Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\cite{luu2023counterfactual,balashankar2023improving} or adversarial
examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
leaves some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

@@

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with
\cite{szegedy2013intriguing}, who define an adversarial attack as an
``imperceptible non-random perturbation''. Thus, we choose to work with
a different distinction between CE and AE than
\cite{freiesleben2022intriguing}, who considers misclassification as the
key distinguishing feature of AE. One of the key observations in this
work is that we can leverage counterfactual explanations during training
and get adversarial examples, essentially for free.

@@

\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values for any of the other penalties in
Equation~\ref{eq-obj}, this effect is not very pronounced.

\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
find that it is not necessary to employ CT during the entire training
phase to achieve the desired improvements in explainability. We have
tested training models conventionally during the first half of training
before switching to CT after this initial ``burn-in'' period and
observed positive results. Put differently, CT may be a way to improve
the explainability of trained models in a fine-tuning manner.

@@

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method benefits from
tuning certain key hyperparameters. In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, counterfactual training is iterative
which makes a variety of methods applicable, including Bayesian
\cite{snoek2012practical} or gradient-based \cite{franceschi2017forward}
optimization.

@@

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; regrettably, all types of ADM solutions without strong external
protections have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}

\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Counterfactual training (CT) combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. For AR recommendations to be actionable, they need to
not only result in plausible counterfactuals but also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. CT aims to improve
model explainability by aligning models with counterfactuals that meet
both desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.,
consistent with the data) and actionable. Formally, we define these
properties as follows:

@@

The characterization of faithfulness and plausibility in
Definition~\ref{def-explainability} is the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search to be
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation}

where \(\text{yloss}(\cdot)\) is a classification loss that induces
discriminative performance (e.g., cross-entropy). The second and third
terms in Equation~\ref{eq-obj} are explained in detail below. For now,
they can be sufficiently described as inducing explainability directly
and indirectly by penalizing: (1) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

\subsection{Directly Inducing Explainability with Contrastive
Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. This follows directly if we
consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
because the JEM objective effectively minimizes the divergence between
the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
generate samples, \cite{grathwohl2020your} rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for initialization
but we depart from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
counterfactual for a randomly sampled factual from a non-target class
generated with a gradient-based CE generator for up to \(T\) iterations.
Mature counterfactuals are ones that have either reached convergence
wrt. the decision threshold \(\tau\) or exhausted \(T\).

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while increasing the
energy of counterfactuals (negative samples) \cite{du2019implicit}. As
the counterfactuals get more plausible
(Definition~\ref{def-explainability}) during training, these opposing
effects gradually balance each other out \cite{lippe2024uvadlc}.

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods facilitate the imposition of
domain and mutability constraints. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition. Like JEMs
\cite{murphy2022probabilistic}, CT can be considered a form of
contrastive representation learning.

\subsection{Indirectly Inducing Explainability with Adversarial
Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\cite{luu2023counterfactual,balashankar2023improving} or AEs
\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
some flexibility with respect to the choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training:

@@

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \cite{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \cite{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations in
this work is that we can leverage CEs during training and get
adversarial examples essentially for free.

@@

\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not very pronounced.

\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
find that it is not necessary to employ CT during the entire training
phase to achieve the desired improvements in explainability. When
training models conventionally during the first 50\% of epochs before
switching to CT for the next 50\% of epochs, we observed positive
results. Put differently, CT may be a way to improve the explainability
of models in a fine-tuning manner.

@@

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method benefits from
tuning certain key hyperparameters. In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

@@

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; all types of ADM solutions without strong external protections
have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

### 2025-03-12 paper/sections/discussion.qmd
Review counterfactual training up to 3.4

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters. In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***Interventions on features may have downstream impacts on fairness.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; regrettably, all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

>>>>

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters. In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***Interventions on features may have downstream impacts on fairness.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

### 2025-03-12 paper/sections/experiments.qmd
Review counterfactual training up to 3.4

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability. We have tested training models conventionally during the first half of training before switching to CT after this initial "burn-in" period and observed positive results. Put differently, CT may be a way to improve the explainability of trained models in a fine-tuning manner.

>>>>

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability. When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.
